---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
---

```{r setoptions, echo=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)

```


## Loading and preprocessing the data
- Package(s) loading
```{r package loading}
library(dplyr)
library(ggplot2)
```

- Data loading
```{r Data Loading}
act_data<- read.csv("activity.csv")
```

- Data Processing
```{r Data Processing }
# Coverting date column from factor to DATE
act_data$date<-as.Date(act_data$date)

# Showing first few rows of the imputed dataset
head(act_data)
```

## What is mean total number of steps taken per day?

- Total number of steps taken per day
```{r total number of steps taken per day }
# Grouping steps count by Date, then adding them accross groups.
totalStepsPerDay<- act_data %>% group_by(date) %>%
    summarise(steps=sum(steps,na.rm=TRUE))
totalStepsPerDay
```

- Histogram of the total number of steps taken each day

```{r Histogram of the total number of steps taken each day }
# Calculating mean and median for the plot
mean1<-round(mean(totalStepsPerDay$steps),2)
median1<-round(median(totalStepsPerDay$steps),2)
# Histogram  generation
ggplot(totalStepsPerDay, aes(steps))+
    geom_histogram(binwidth = 500,alpha=0.7, col="black",aes( fill=..count..,))+
    geom_density(aes(y=..count.. *500))+
    geom_vline(xintercept = mean1,color="red",size=1.5,alpha=0.5)+
    geom_text(aes(x=8000, label=paste("Mean\n",mean1), y=8.5))+
    geom_vline(xintercept = median1,color="darkgreen",size=1.5,alpha=0.5)+
    geom_text(aes(x=12000, label=paste("Median\n",median1), y=8.5))+
    labs(title = "Histogram: Total number of steps taken each day", x="Daily steps", y="Frequency")+
    theme(legend.position = c(0.95,0.85), plot.title = element_text(hjust = 0.5))
```

- Mean and median of the total number of steps taken per day
```{r Mean and median of the total number of steps taken per day }
# Mean calculation
mean1<-round(mean(totalStepsPerDay$steps),2)
print(paste("Mean = ",mean1))

# Median calculation
median1<-round(median(totalStepsPerDay$steps),2)
print(paste("Median = ",median1))
```

## What is the average daily activity pattern?
- Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r time series plor }
# Grouping steps count by interval, then calculating means across groups
timeSeriesData<- act_data %>% group_by(interval) %>% summarise(steps=mean(steps,na.rm=TRUE))
# Time series plot
ggplot(timeSeriesData, aes(x=interval, y=steps))+
    geom_line(color="red")+
    labs(title = "Time series plot: Average number of steps taken by 5 Minute interval ",
         x="Five minute interval Identifier", y="Steps count averaged over all days")+
    theme( plot.title = element_text(hjust = 0.5))
```

- 5-minute interval with the maximum number of steps (average across all the days in the dataset)
```{r maxinterval }
maxinterval<- timeSeriesData[which.max(timeSeriesData$steps),]
names(maxinterval)<- c("Interval identifier", "Maximun average steps")
maxinterval
```

## Imputing missing values
-  Total number of missing values in the original dataset 
```{r missing values }
colSums(is.na(act_data))
```

- Replacing missing values with mean of steps across all days for that specific 5 minute interval
```{r missing value replacement }
# Grouping steps count by interval, then replacing missing steps count value with mean of steps across all days for that specific 5 minute interval
imputed_data<- act_data %>% group_by(interval) %>% 
    mutate(steps= replace(steps, is.na(steps), mean(steps,na.rm=TRUE) ))
# Checking missing value number in the new formatted dataset
colSums(is.na(imputed_data))
```

- New imputed dataset
```{r imputed dataset }
# Showing first few rows of the imputed dataset
head(imputed_data)
```

- Imputed data - Histogram, mean and median
```{r Imputed data - Histogram, mean and median }
# Grouping steps count by Date, Then adding steps count across groups
totalStepsPerDay_imputed<- imputed_data %>%
    group_by(date) %>%
    summarise(steps=sum(steps,na.rm=TRUE))
# calculating mean and median for the histogram
mean2<-round(mean(totalStepsPerDay_imputed$steps),2)
median2<-round(median(totalStepsPerDay_imputed$steps),2)
# Histogram  generation
ggplot(totalStepsPerDay_imputed, aes(steps))+
    geom_histogram(binwidth = 500,alpha=0.7, col="black",aes( fill=..count..,))+
    geom_density(aes(y=..count.. *500))+
    geom_vline(xintercept = mean2,color="red",size=1.5,alpha=0.5)+
    geom_text(aes(x=8000, label=paste("Mean\n",mean2), y=8.5))+
    geom_vline(xintercept = median2,color="darkgreen",size=1.5,alpha=0.5)+
    geom_text(aes(x=13000, label=paste("Median\n",median2), y=8.5))+
    labs(title = "Histogram: Total number of steps taken each day", x="Daily steps", y="Frequency")+
    theme(legend.position = c(0.95,0.85), plot.title = element_text(hjust = 0.5))
```

THe mean and median value is same. That is why, in the histogram, the red mean line and green median line merged and created a brown line.  

```{r mean and median }
# Mean 
print(paste("Mean = ",mean2))

# Median
print(paste("Median = ",median2))
```

According to the histogram and the calculated mean and median values, Imputing missing values in the dataset made the steps count variable more normalized. Also, the mean and median value became same.

## Are there differences in activity patterns between weekdays and weekends?
- New factor variable in the dataset with two levels – “weekday” and “weekend” 
```{r adding new factor weekday type}
# Adding new column to specify the weekdays based on the dates
imputed_data$weekday <- weekdays(imputed_data$date)

# Adding new column named weekdayType to specify whether the date is a weekday or a weekend
imputed_data$weekdayType<-ifelse(weekdays(imputed_data$date) %in% c("Saturday","Sunday"), "Weekend", "Weekday")

# Showing first few rows of the modified data 
head(imputed_data)

```

- Panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r panel plot}

# Grouping steps count by weekdayType and interval accordingly, Then mean of steps count across groups
timeSeriesData2<- imputed_data %>% group_by(weekdayType,interval) %>% summarise(steps = mean(steps, na.rm=TRUE))

# Panel plot
ggplot(timeSeriesData2, aes(x=interval, y=steps, color=weekdayType))+
    facet_wrap(weekdayType~.,ncol = 1)+
    geom_line()+
    labs(title = "Time series plot: Average number of steps taken by 5 Minute interval ",
         x="Five minute interval Identifier", y="Steps count averaged over all days")+
    theme( plot.title = element_text(hjust = 0.5))
```

